sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-09 doc/isec_abstract.Rmd
Edits

---
title: "ISEC_abstract"
output: pdf_document
---

@@

Analyzing geostatistical data (coordinate referenced observations) is becoming increasingly common in ecology; these data often include a discrete temporal dimension, such as monthly or yearly samples. Computational tools to model spatial and spatiotemporal processes have advanced rapidly over the last decade; examples of R packages for analyzing geostatistical data include `mgcv`, `INLA` and `inlabru`, `VAST`, and `spBayes`. These packages are 
In this talk, we present a new R package `sdmTMB`. The goals of `sdmTMB` are not to replace functionality in existing packages, but provide a user friendly implementation of SPDE models -- similar to the popular package `glmmTMB`, but with the addition of spatial models. We extend the generalized linear mixed models (GLMMs) familiar to ecologists to include the following optional features:
* spatial fields
* spatiotemporal fields, that may be independent by year or modeled with random walks or autoregressive processes
* smooth terms for covariates, using the familiar `s()` notation in `mgcv`
* spatially varying coefficient models (SVC)
* non-stationary covariates (coefficients modeled as random walks)

Like `glmmTMB`, estimation is performed in `sdmTMB` via maximum likelihood, implemented  using the Template Model Builder (TMB) software.

We provide a general overview of the package structure, and demonstrate features using several examples (Snowy Owl distribution in North America, and a dataset describing the distribution of Pacific cod in British Columbia).

>>>>

---
title: "ISEC_abstract"
output: word_document
---

@@

Analyzing geostatistical data (coordinate-referenced observations) is becoming increasingly common in ecology; these data often include a discrete temporal dimension, such as monthly or yearly samples. Computational tools to model spatial and spatiotemporal processes have advanced rapidly over the last decade; examples of R packages for analyzing geostatistical data include mgcv, INLA and inlabru, VAST, and spBayes. 
SOMETHING HERE...
In this talk, we present a new R package sdmTMB. The goals of sdmTMB are not to replace functionality in existing packages, but provide a user-friendly implementation of SPDE models---similar to the popular R package glmmTMB, but with the addition of spatial and spatiotemporal models with an SPDE approach.
We extend the generalized linear mixed models (GLMMs) familiar to ecologists to include the following optional features:

* spatial fields
* spatiotemporal fields that may be independent by year or modeled with random walks or autoregressive processes
* smooth terms for covariates, using the familiar `s()` notation in mgcv
* spatially varying coefficient models (SVCs)
* non-stationary covariates (coefficients modeled as random walks)

Estimation is performed in sdmTMB via maximum marginal likelihood, implemented using Template Model Builder (TMB).
The sdmTMB package also allows for models to be passed to Stan, allowing for efficient Bayesian estimation.
We provide a general overview of the package structure, and demonstrate features using several examples including Snowy Owl winter irruptions in North America and a dataset describing the distribution of Pacific Cod in British Columbia. 
We highlight advantages of using sdmTMB for these classes of models including (1) estimation in TMB is often faster than alternative estimation routines, (2) sdmTMB provides a user-friendly interface that is familiar to many R users, (3) sdmTMB allows for easy out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficient models). 
For analysts fitting species distribution models (SDMs) or estimating population abundance trends, sdmTMB provides specific functions for performing these calculations.

### 2021-12-09 doc/paper.Rmd
Edits

@@

Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
Future development may include zero-inflated models, improvements to Bayesian sampling efficiency, continuous time models (REF?), and non-gaussian random fields [e.g., @anderson2019].
<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend. Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data.

>>>>

We could generate a population index that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, APPENDIX).

@@

Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, spatial covariance barriers, time-varying spatiotemporal variance, and simulation from the parameter joint precision matrix.
Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-gaussian random fields [e.g., @anderson2019].
<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend. Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data.
